---
title: Heat and Storms in the United States Negatively Impact Human Life and Cost
  Billions in Property and Food Loss
author: "sdbj2063"
date: "Sunday, January 25, 2015"
output: pdf_document
keep_md: yes
---

Heat and Storms in the United States Negatively 
===============================================
Impact Human Life and 
=======================
Cost Billions in Property and Food Loss
=======================================
### Johns Hopkins and Coursera
### Reproducible Research Class
### Peer Assessment 2

#### January 24, 2015

#### user sdbj2063

# Synopsis (10 sentences max.)

By calculating the mean for injuries, fatalities, property damage and crop damage based upon the severe 
weather event type, it became obvious that global warming is having an adverse impact on 
human health, property and food sources.

The project used the data set "Storm Data" from the National Oceanic and Atmospheric Administration (NOAA) 
for 1950 through November 2011. 

The project required developers to determine which types of storm weather events in the United States 
were the most harmful to "population health" and which types had the "greatest economic consequences."

Regarding the impact on human life, more event types had injuries than deaths, and the largest event type 
for injuries had almost triple the number of incidents compared to the largest event type for fatalities. 

Eleven event types had both injuries and fatalities of more than one person, and heat-related events 
accounted for a high number of fatalities and injuries.

Regarding economic impact, more event types had property damage than crop damage. 

While the largest event type based upon dollars for property damage was greater than the largest event 
for crop damage, both involved rain or moisture.

Property damage and crop damage occured jointly in 111 event types, and storm-related events accounted 
for a high number of property and crop damage events.

Finally, eight events had all four factors--human injuries and fatalities, property and crop damage--
and five of those events were directly related to heat.


# Documentation Conventions

In the **Result** section, each project problem has a heading **Problem**.  

Beneath that are **Questions** that the script uses to address the problem, **CONCLUSIONS** about the 
data processed for the problem with script and explanation and an **Illustrative Graph** for each problem.

R functions appear in the text blocks the first time as *function-name* in *italics* followed by 
"function." Subsequent references use the format *function-name()*.

Calculated variable values are displayed within the code block results and within text blocks as 
`variable-name` "variable."  

Field or variable names within an R data frame appear within the text blocks as **field-name** 
in bold followed by "field."



# Data Processing

## Loading the Data

Several libraries were necessary to complete the aggregate computations and develop the boxplots.

```{r dploadlibraries, echo=TRUE}
## Set up libraries

library(base)
library(utils)
library(graphics)
library(grDevices)
library(stringr)
library(plyr)
library(lattice)
```

The script downloads the *.bz2 archive and loads the data into a data frame, eliminating 
unnecessary columns.  

```{r loaddata, echo=TRUE}
## Download the zip file and overwrite an existing file
myfile <- "repdata_data_StormData.csv.bz2"
mypath <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

download.file(mypath, myfile, quiet = FALSE, mode = "wb")
```

The usable fields included the event type, **EVTYPE** field; counts for injuries and fatalities, 
**INJURIES** and **FATALITIES** fields; and total dollar amounts per event for property 
damage and crop damage, **PROPDMG**, **PROPDMGEXP**, **CROPDMG** and **CROPDMGEXP** fields.  
The two **EXP** fields are the indicators for the demonation of the numeric value, such as "K" 
for thousands. **EVTYPE** became a character string to ease searching.  

```{r definecolumns, echo=TRUE}
## Unzip the file and load seven data columns into R
mycolclasses <- c(rep("NULL", 7), "factor", rep("NULL", 14), rep("numeric", 3), "factor", "numeric", "factor", rep("NULL", 9))

mydata <- read.csv(bzfile(myfile), colClasses = mycolclasses)

## make EVTYPE character for string searching purposes
mydata$EVTYPE <- as.character(mydata$EVTYPE)
```

The resulting format is small and better prepared for manipulation.

```{r premydata, echo=TRUE}
str(mydata)
```

## Preprocessing Property and Crop Damage Data

To standardize the approach to answering the questions, I chose to find the average of the values fields 
for each event type using the **EVTYPE** field. While the raw data was suitable for human fatalities and 
injuries, the property and crop damage figures needed preprocessing to achieve a usable number.

The original data set contained two fields for the property damage, a numerical figure **PROPDMG** and a 
character field **PROPDMGEXP** indicating what U.S. Dollar amount should be included. According to the 
NATIONAL WEATHER SERVICE INSTRUCTION 10-1605, "STORM DATA PREPARATION,"  "K" stood for 
thousands of dollars, "M" stood for millions and "B" stood for billions. The character field contained 
other values the project documentation did not acknowledge, so dollar calculations ignored those values.   

```{r p2setup, echo=TRUE}
## Obtain list of demoninator values for property and crop damage
mydamage.property.u <- as.character(unique(mydata$PROPDMGEXP))
mydamage.crop.u <- as.character(unique(mydata$CROPDMGEXP))

mydamage.property.u
mydamage.crop.u 


```

Once the script identified the denominator factors, it added two fields to the original data set, one for 
the calculated the total dollar cost for property damage and the second for crop damage.


```{r dpaddtotalcostfields, echo=TRUE}
## Add new calculated total cost fields for property damage and crop damage.
mydata$PROPDMGCOST <- 0
mydata$CROPDMGCOST <- 0    
```

By multiplying the property number **PROPDMG** by the appropriate operator based upon the denominator in 
**PROPDMGEXP** the script calculated the total dollar amount for each event.

```{r p1calctotalcostproperty, echo=TRUE}
## Calculate total cost PROPDMGCOST
## Use three values identified in paperwork, K, M and B
## Multiple PROPDMG > 0 by appropriate valuation

## Thousands using K
mydata[mydata$PROPDMGEXP %in% c(mydamage.property.u[1]) & mydata$PROPDMG > 0, 8 ] <- as.numeric(mydata[mydata$PROPDMGEXP %in% c
(mydamage.property.u[1]) & mydata$PROPDMG > 0, 4 ] * 1000)

## Millions using M
mydata[mydata$PROPDMGEXP %in% c(mydamage.property.u[2]) & mydata$PROPDMG > 0, 8 ] <- as.numeric(mydata[mydata$PROPDMGEXP %in% c
(mydamage.property.u[2]) & mydata$PROPDMG > 0, 4 ] * 1000000)

## Billions using B
mydata[mydata$PROPDMGEXP %in% c(mydamage.property.u[4]) & mydata$PROPDMG > 0, 8 ] <- as.numeric(mydata[mydata$PROPDMGEXP %in% c
(mydamage.property.u[4]) & mydata$PROPDMG > 0, 4 ] * 1000000000)
```


The original data set contained two fields for the crop damage, a numerical figure **CROPDMG** and a 
character field **CROPDMGEXP** indicating what U.S. Dollar amount should be included. According to the 
NATIONAL WEATHER SERVICE INSTRUCTION 10-1605, "STORM DATA PREPARATION,"  "K" stood for 
thousands of dollars, "M" stood for millions and "B" stood for billions. The character field contained 
other values the project documentation did not acknowledge, so dollar calculations ignored those values.   

By multiplying the property number **CROPDMG** by the appropriate operator based upon the denominator in 
**CROPDMGEXP** the script calculated the total dollar amount for each event.

```{r p1calctotalcostcrop, echo=TRUE}
## Calculate total cost CROPDMGCOST
## Use three values identified in paperwork, K, M and B
## Multiple CROPDMG > 0 by appropriate valuation

## Thousands using K
mydata[mydata$CROPDMGEXP %in% c(mydamage.crop.u[3]) & mydata$CROPDMG > 0, 9 ] <- as.numeric(mydata[mydata$CROPDMGEXP %in% c
(mydamage.crop.u[3]) & mydata$CROPDMG > 0, 6 ] * 1000)

## Millions using M
mydata[mydata$CROPDMGEXP %in% c(mydamage.crop.u[2]) & mydata$CROPDMG > 0, 9 ] <- as.numeric(mydata[mydata$CROPDMGEXP %in% c
(mydamage.crop.u[2]) & mydata$CROPDMG > 0, 6 ] * 1000000)

## Billions using B
mydata[mydata$CROPDMGEXP %in% c(mydamage.crop.u[5]) & mydata$CROPDMG > 0, 9 ] <- as.numeric(mydata[mydata$CROPDMGEXP %in% c
(mydamage.crop.u[5]) & mydata$CROPDMG > 0, 6 ] * 1000000000)

```

The final step created a new data set with damage values greater than zero (0).

```{r p2totalpropertycrop, echo=TRUE}
## Create new data set. 
## Keep rows with property or crop damage amounts greater than 0 to reduce the size of the data set.

mydata.2 <- mydata[mydata$PROPDMGCOST > 0 | mydata$CROPDMGCOST > 0, c(1,8,9)]
```

More data calculations and instructions pertaining to each question and the strategy to reach conclusions  
are in the **Results** section of this document.  

# Results

The results are divided into two sections, one for each of the problems presented in the project. 
Each section has one two-boxplot layout to show the range of impact for each category--
human injuries and fatalities and damage to property and crops. The final conclusion examines the entire 
data set of mean values to identify what event types have all four categories--injuries 
and fatalities for humans, property damage and crop damage.  

## Summary of Results

I used the mean, or average, of each of the four key numeric fields--number of fatalities and number of 
injuries for human impact and the total cost of property damage and total cost of crop 
damage for economic impact--grouped by the type of event to reach basic conclusions about the impact 
of severe weather events on the United States of America's population and economy.  

While all the events could have a potential economic impact for human beings, I focused on direct 
affects to life and limb for humans.  

Pertaining to events harmful to human population health:  
1. More event types resulted in injuries to human beings than deaths.  
2. The largest event type for injuries was almost triple the number of incidents for the largest 
event type for fatalities.  
3. Eleven event types had both injuries and fatalities of more than one person.  
4. Heat-related events accounted for a high number of both fatalities and injuries.  

Pertaining to events having the greatest economic impact:  
1. More event types reesult in property damage than crop damage.  
2. The largest event type for property damage has more damage than the largest event for crop damage, 
but both involve rain or moisture.  
3. In total, 111 event types have both property damage and crop damage.  
4. Storm-related events accounted for a high number of events with property or crop damage.  

Concluding finding:  
Eight events had all four factors--human injuries and fatalities, property and crop damage--
and five of those events were directly related to heat.  

## Problem One:  Across the U.S., which event types were most harmful to population health?  

The original data set recorded two variables with data pertinent to human health--injuries and 
fatalities. To address this problem, I focused on calculating the mean for each event type for 
injuries and fatalities using the **INJURIES** field, **FATALITIES** field and **EVTYPE** field.  


### QUESTION: What was the mean number of injuries to humans based upon event?
 
The strategy was to calculate the mean number of injuries for each event type and limit the data set 
to events with more than one occurrence. Original examination of the data showed a plethera of events 
with one occurrence. The final step was to order the data in descending order based upon the number 
of injuries.  

```{r p1meaninjuries, echo=TRUE}
## Caculate mean amounts using the aggregate function and group by event type.
mydata.mean.injuries <- aggregate(INJURIES ~ EVTYPE, data = mydata, mean)

## Keep rows that do not equal zero.
mydata.mean.injuries.1 <- mydata.mean.injuries[mydata.mean.injuries$INJURIES > 1,]

## Order the data in descending order based upon the mean number injuries per event type.
mydata.mean.injuries.1 <- mydata.mean.injuries.1[with(mydata.mean.injuries.1, order(-INJURIES)), ]
```

The resulting data set showed that the mean number of injuries was no more than 70 and most were below 10.  

```{r p1meaninjurieslist, echo=TRUE}
## List event types in descending order based upon the mean number of injuries
mydata.mean.injuries.1

```

### QUESTION: What was the mean number of fatalities for humans based upon event?   

The strategy was to calculate the mean number of fatalities for each event type and limit the data set 
to events with more than one occurrence. Original examination of the data showed a plethera of events 
with one occurrence. The final step was to order the data in descending order based upon the number 
of fatalities.  

```{r p1meanfatalities, echo=TRUE}
## Caculate mean amounts using the aggregate function and group by event type.
mydata.mean.fatalities <- aggregate(FATALITIES ~ EVTYPE, data = mydata, mean)

## Keep rows that do not equal zero.
mydata.mean.fatalities.1 <- mydata.mean.fatalities[mydata.mean.fatalities$FATALITIES > 1,]

## Order the data in descending order based upon the mean number of fatalities per event type.
mydata.mean.fatalities.1 <- mydata.mean.fatalities.1[with(mydata.mean.fatalities.1, order(-FATALITIES)), ]
```

The resulting data set showed that most of the events had a mean number of fatalities below 10.  

```{r p1meanfatalitieslist, echo=TRUE}
## List event types in descending order based upon the mean number of fatalities
mydata.mean.fatalities.1
```

## CONCLUSION #1

More event types resulted in injuries to human beings than deaths.  

The script counted the number of rows in the respective data sets for injuries and fatalities 
to obtain the total number of event types for each indicent category.  

```{r p1c1injuriesfatalities, echo=TRUE}
## Record the number of event types for injuries
mydata.mean.injuries.1.no <- nrow(mydata.mean.injuries.1)
mydata.mean.injuries.1.no

## Record the number of event types for fatalities
mydata.mean.fatalities.1.no <- nrow(mydata.mean.fatalities.1)
mydata.mean.fatalities.1.no

```



## CONCLUSION #2

The largest event type for injuries was almost triple the number of incidents for the largest 
event type for fatalities.   

The script collected the first value in the ordered data sets for injuries and fatalities, respectively.  

```{r p1highestnumber, echo=TRUE}
## Record the highest injury event type
mydata.mean.injuries.1.high <- mydata.mean.injuries.1[1,]
mydata.mean.injuries.1.high

## Record the highest fatality event type
mydata.mean.fatalities.1.high <- mydata.mean.fatalities.1[1,]
mydata.mean.fatalities.1.high

```
## Illustrative Graph

The side-by-side boxplot graphs showed the mean number of injuries and fatalities on the same scale. 
The graph marked the box boundaries for the 1st through 3rd quartiles. Outliers appeared as circles.  

```{r p1plotinjuriesfatalities, echo=TRUE}
## Set graph properties for property and crop damage
mymain.injuries <- "Mean Number of Injuries"
myxlab.injuries <- "1st Quartile to 3rd Quartile Box"
myylab.injuries <- "Mean Number"
myylim.injuries <- c(0,mydata.mean.injuries.1.high[1,2])
mymain.fatalities <- "Mean Number of Fatalities"
myxlab.fatalities <- "1st Quartile to 3rd Quartile Box"
myylab.fatalities <- "Mean Number"
myylim.fatalities <- c(0,mydata.mean.injuries.1.high[1,2])

## Create one row, two column layout for the injuries and fatalities boxplots, respectively
par(mfrow = c(1, 2))

boxplot(mydata.mean.injuries.1$INJURIES, main=mymain.injuries, xlab=myxlab.injuries, ylab=myylab.injuries, ylim=myylim.injuries)

boxplot(mydata.mean.fatalities.1$FATALITIES, main=mymain.fatalities, xlab=myxlab.fatalities, ylab=myylab.fatalities, 
ylim=myylim.fatalities)

```


## CONCLUSION #3

Eleven event types had both injuries and fatalities of more than one person.  

The script merged the injury and fatality data sets, focusing on common event types. It calculated 
the number of events by counting the number of rows in the merged data set.   

```{r p1mergedata, echo=TRUE}
## Merge the injury and fatality mean data sets and order in descending order by injuries and fatalities 
mydata.mean.merge.people <- merge(mydata.mean.injuries.1, mydata.mean.fatalities.1, by.x="EVTYPE", by.y="EVTYPE")
mydata.mean.merge.people <- mydata.mean.merge.people[with(mydata.mean.merge.people, order(-INJURIES, -FATALITIES)), ]
mydata.mean.merge.people

## Count the number of rows to calculate the number of event types with both injuries and fatalities
mydata.mean.merge.people.no <- nrow(mydata.mean.merge.people)
mydata.mean.merge.people.no
```

## CONCLUSION #4

Heat-related events accounted for a high number of fatalities and injuries.  

The script searched the event types for "heat" and calculated the number of events for injuries 
and fatalities by counting the number of rows that met the criteria.  

```{r p1meanheat, echo=TRUE}
## Search for injury event types with "heat"
mydata.mean.injuries.heat <- grep("[Hh][Ee][Aa][Tt]", mydata.mean.injuries.1$EVTYPE, value=TRUE)
mydata.mean.injuries.heat

## Calculate the number of events for injuries with "heat" in the name             
mydata.mean.injuries.heat.no <- length(grep("[Hh][Ee][Aa][Tt]", mydata.mean.injuries.1$EVTYPE, value=TRUE))
mydata.mean.injuries.heat.no
## Count of total injuries
mydata.mean.injuries.1.no

## Search for fatality event types with "heat"
mydata.mean.fatalities.heat <- grep("[Hh][Ee][Aa][Tt]", mydata.mean.fatalities.1$EVTYPE, value=TRUE)
mydata.mean.fatalities.heat

## Calculate the number of events for fatalities with "heat" in the name
mydata.mean.fatalities.heat.no <- length(grep("[Hh][Ee][Aa][Tt]", mydata.mean.fatalities.1$EVTYPE, value=TRUE))
mydata.mean.fatalities.heat.no
## Count of total fatalities
mydata.mean.fatalities.1.no

```

Out of the events with injuries, `mydata.mean.injuries.1.no` variable with 
**`r format(mydata.mean.injuries.1.no, 2)`**  events, **`r format(mydata.mean.injuries.heat.no,2)`** events were 
related to heat, based upon the `mydata.mean.injuries.heat.no` variable.   

Out of the events with fatalities, `mydata.mean.fatalities.1.no` variable with 
**`r format(mydata.mean.fatalities.1.no, 2)`**  events, **`r format(mydata.mean.fatalities.heat.no, 2)`** events were 
related to heat, based upon the `mydata.mean.fatalities.heat.no` variable.   

## Problem 2: Across the U.S., which types of events had the greatest economic consequences?

To address this problem, I focused on calculating the mean for each event type for property damage 
and crop damage. The numbers were so large as to be meaningless in a graph, so the illustrative 
graph uses the log scale.  



### QUESTION: What was the mean dollar amount of property damage based upon event type?   

The strategy was to calculate the mean amount for each event type and limit the data set to 
events with more than zero occurrences. The final step was to order the data in descending order 
based upon dollar amount.  


```{r p2calcmeanproperty, echo=TRUE}
## Caculate mean amounts using the aggregate function and group by event type.
mydata.mean.property <- aggregate(PROPDMGCOST ~ EVTYPE, data = mydata.2, mean)

## Keep rows that do not equal zero.
mydata.mean.property.1 <- mydata.mean.property[mydata.mean.property$PROPDMGCOST != 0,]

## Order the data in descending order based upon the total cost per event type.
mydata.mean.property.1 <- mydata.mean.property.1[with(mydata.mean.property.1, order(-PROPDMGCOST)), ]
```

The resulting data set, below, ordered from largest dollar amount descending, was exhausive. Further processing was necessary.   

```{r p2meanpropertylist, echo=TRUE}
mydata.mean.property.1
```


### QUESTION: What was the mean dollar amount of crop damage based upon event type?   
 
The strategy was to calculate the mean amount for each event type and limit the data set to 
events with more than zero occurrences. The final step was to order the data in descending 
order based upon dollar amount.  


```{r p2calcmeancrop, echo=TRUE}
## Caculate mean amounts using the aggregate function and group by event type.
mydata.mean.crop <- aggregate(CROPDMGCOST ~ EVTYPE, data = mydata.2, mean)

## Keep rows that do not equal zero.
mydata.mean.crop.1 <- mydata.mean.crop[mydata.mean.crop$CROPDMGCOST != 0,]

## Order the data in descending order based upon the total cost per event type.
mydata.mean.crop.1 <- mydata.mean.crop.1[with(mydata.mean.crop.1, order(-CROPDMGCOST)), ]
```

The resulting data set, below, ordered from largest dollar amount descending, was exhausive. Further processing was necessary.   

```{r p2meancroplist, echo=TRUE}
mydata.mean.crop.1
```


## CONCLUSION #1

More event types resulted in property damage than crop damage.  

The script counted the number of rows in the respective data sets for property and crop damage 
to obtain the total number of event types for each damage category.  

```{r p2meanpropertycropno, echo=TRUE}
## Calculate the total number of event types with property damage
mydata.mean.property.1.no <- nrow(mydata.mean.property.1)
mydata.mean.property.1.no


## Calculate the total number of event types with crop damage
mydata.mean.crop.1.no <- nrow(mydata.mean.crop.1)
mydata.mean.crop.1.no

```



## CONCLUSION #2

The largest event type for property damage had more damage than the largest event for crop damage, 
but both involved rain or moisture.   

The script collected the first value in the ordered data sets for property and crop damage, respectively.  


```{r p2propertycrophigh, echo=TRUE}
## Identify the event type with the highest property damage
mydata.mean.property.1.high <- mydata.mean.property.1[1,]
mydata.mean.property.1.high


## Identify the event type with the highest crop damage
mydata.mean.crop.1.high <- mydata.mean.crop.1[1,]
mydata.mean.crop.1.high

```

## Illustrative Graph

The side-by-side boxplot graphs showed the mean dollar amount of property and crop damage on 
the same log scale. Due to the size of the dollar amounts, the y-axis scale used the log of the 
mean values. The graph marked the box boundaries for the 1st and 3rd quartiles. Outliers appeared as 
circles.


```{r p2plotboxplot, echo=TRUE}
## Set graph properties for property and crop damage
mymain.property <- "Mean $ Amount Property Damage"
myxlab.property <- "1st Quartile to 3rd Quartile Box"
myylab.property <- "Mean Number on Log Scale"
myylim.property <- c(0,log(mydata.mean.property.1.high[1,2]))
mymain.crop <- "Mean $ Amount Crop Damage"
myxlab.crop <- "1st Quartile to 3rd Quartile Box"
myylab.crop <- "Mean Number on Log Scale"
myylim.crop <- c(0,log(mydata.mean.property.1.high[1,2]))

## Create one row, two column layout for the property and crop damage boxplots, respectively
par(mfrow = c(1, 2))

boxplot(log(mydata.mean.property.1$PROPDMGCOST), main=mymain.property, xlab=myxlab.property, ylab=myylab.property, 
ylim=myylim.property)

boxplot(log(mydata.mean.crop.1$CROPDMGCOST), main=mymain.crop, xlab=myxlab.crop, ylab=myylab.crop, ylim=myylim.crop)

```

## CONCLUSION #3

In total, 111 event types had both property damage and crop damage.  

The script merged the property damage and crop damage data sets, focusing on common event types. 
It calculated the number of events by counting the number of rows in the merged data set.  


```{r p2mergepropertycrop, echo=TRUE}
## Merge the property and crop mean data sets and order in descending order by property damage and crop damage
mydata.mean.merge.nonpeople <- merge(mydata.mean.property.1, mydata.mean.crop.1, by.x="EVTYPE", by.y="EVTYPE")
mydata.mean.merge.nonpeople <- mydata.mean.merge.nonpeople[with(mydata.mean.merge.nonpeople, order(-PROPDMGCOST, -CROPDMGCOST)), ]
mydata.mean.merge.nonpeople

## Count the number of rows to calculate the number of event types with both property and crop damage
mydata.mean.merge.nonpeople.no <- nrow(mydata.mean.merge.nonpeople)
mydata.mean.merge.nonpeople.no
```
 

## CONCLUSION #4

Storm-related events accounted for many of the events relating to property and crop damage.  

The original data set was titled "Storm Data," but most of the possible event descriptions 
did not relate specifically to storms.  To see whether any "storms" impacted property or crop 
damage I searched the event names on each list of events with calculated means of damage.

The script searched the event types for "storm" and calculated the number of events for property 
and crop damage by counting the number of rows that "storm" in the name.  


```{r p2stormdamage, echo=TRUE}
## Search for property damage event types with "storm"
mydata.mean.property.storm <- grep("[Ss][Tt][Oo][Rr][Mm]", mydata.mean.property.1$EVTYPE, value=TRUE)
mydata.mean.property.storm

## Calculate the number of events for property damage with "storm" in the name
mydata.mean.property.storm.no <- length(grep("[Ss][Tt][Oo][Rr][Mm]", mydata.mean.property.1$EVTYPE, value=TRUE))
mydata.mean.property.storm.no
## count of total property events
mydata.mean.property.1.no

## Search for crop damage event types with "storm"
mydata.mean.crop.storm <- grep("[Ss][Tt][Oo][Rr][Mm]", mydata.mean.crop.1$EVTYPE, value=TRUE)
mydata.mean.crop.storm
             
## Calculate the number of events for crop damage with "storm" in the name
mydata.mean.crop.storm.no <- length(grep("[Ss][Tt][Oo][Rr][Mm]", mydata.mean.crop.1$EVTYPE, value=TRUE))
mydata.mean.crop.storm.no
## count of total crop events
mydata.mean.crop.1.no
```

Out of the events with property damage, `mydata.mean.property.1.no` variable with 
**`r format(mydata.mean.property.1.no,2)`**  events, **`r format(mydata.mean.property.storm.no,2)`** events were 
related to storms, based upon the `mydata.mean.property.storm.no` variable.   

Out of the events with crop damage, `mydata.mean.crop.1.no` variable with 
**`r format(mydata.mean.crop.1.no, 2)`**  events, `r format(mydata.mean.crop.storm.no, 2)`** events were 
related to storms, based upon the `mydata.mean.crop.storm.no` variable.   

## FINAL CONCLUSION

Eight events had all four factors--human injuries and fatalities, property and crop damage--
and five of those events were directly related to heat.

The script merged the four mean data sets, looking for event types common to all four data sets.


```{r finalmerge, echo=TRUE}
## Merge four mean data sets into one, with matching on the "event type" EVTYPE field.

mydata.mean.merge.all <- merge(mydata.mean.injuries.1, mydata.mean.fatalities.1, by.x="EVTYPE", by.y="EVTYPE")
mydata.mean.merge.all <- merge(mydata.mean.merge.all, mydata.mean.property.1, by.x="EVTYPE", by.y="EVTYPE")
mydata.mean.merge.all <- merge(mydata.mean.merge.all, mydata.mean.crop.1, by.x="EVTYPE", by.y="EVTYPE")
mydata.mean.merge.all

## Calculate the number of rows of the merged data set to obtain the total count
mydata.mean.merge.all.no <- nrow(mydata.mean.merge.all)
mydata.mean.merge.all.no
```

The script then determined the number of event types that contained "heat."

```{r finalheatcalc, echo=TRUE}
## Calculate which event types have "heat" in their names
mydata.mean.merge.all.heat <- grep("[Hh][Ee][Aa][Tt]", mydata.mean.merge.all$EVTYPE, value=TRUE)
mydata.mean.merge.all.heat

## Calculate the number of event types with "heat" in their names
mydata.mean.merge.all.heat.no <- length(grep("[Hh][Ee][Aa][Tt]", mydata.mean.merge.all$EVTYPE, value=TRUE))
mydata.mean.merge.all.heat.no
```


# System Configuration

The following version of R and computer platform generated the code and graphs.

```{r versioninfo, echo=TRUE}
## Version information
myrversion <- R.Version()
myrversion$version.string
myrversion$platform
```

The internet connection speed allows sufficient time to download the file. The most time intensive 
part of the process is loading the data. Limiting the number of columns, or variables, significantly 
reduced the run-time speed.

# Appendix I:     Data Definitions

### Data definition
#### Loaded data with crop and property damage calculations.

```{r ddmydata, echo=TRUE}
str(mydata)
```

### Data definition
#### Merged injury and fatality data with common event types

```{r ddmeanmergepeople, echo=TRUE}
str(mydata.mean.merge.people)
```


### Data definition
#### Merged crop and property damage data with common event types

```{r ddmeanmergenonpeople, echo=TRUE}
str(mydata.mean.merge.nonpeople)
```

### Data definition
#### Merged data for human fatalities and injuries and property and crop damage with common event types

```{r ddmeanmergeall, echo=TRUE}
str(mydata.mean.merge.all)
```

# Appendix II:	Other References


"NOAA National Severe Storms Laboratory." NSSL: Severe Weather 101. Accessed January 25, 2015. http://www.nssl.noaa.gov/education/svrwx101/.

The web site lists seven severe storm types:  thunderstorms, tornadoes, floods, lightening, hail, 
damaging winds and winter weather.

